---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  dpi = 300
)
```

# US COVID-19 cases by 2016 election vote

<!-- badges: start -->
<!-- badges: end -->

The *Washington Post* published an article by Philip Bump on June 17, titled ["Coronavirus has come to Trump country"](https://www.washingtonpost.com/politics/2020/06/17/coronavirus-has-come-trump-country/). The article included a striking figure showing daily COVID-19 cases over time, aggregated by 2016 election results (i.e. whether a county or state voted for Donald Trump or Hillary Clinton).

The short script below provides R code for reproducing this figure with up-to-date data. The script was run on **`r Sys.Date()`**, but it will automatically pull in the most recent data whenever you run it.

```{r covote, warning=FALSE}
library(data.table)
library(ggplot2)
library(hrbrthemes)

# COVID-19 data -----------------------------------------------------------

## Source: NY Times COVID-19 data repo: https://github.com/nytimes/covid-19-data
nytc = fread('https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv')
nytc[, `:=` (date = as.Date(date), fips = sprintf('%05d', fips))]

# 2016 election data ------------------------------------------------------

## Counties (Source: https://github.com/tonmcg/US_County_Level_Election_Results_08-16)
us2016_counties = fread('https://raw.githubusercontent.com/tonmcg/US_County_Level_Election_Results_08-16/master/2016_US_County_Level_Presidential_Results.csv')
us2016_counties =
  us2016_counties[, `:=` (fips = sprintf('%05d', combined_fips),
                          result_county = fifelse(per_dem > per_gop, 'Clinton won', 'Trump won'))] %>%
  .[, .(fips, state_abbr, result_county)]

## States
us2016_states = fread('https://vincentarelbundock.github.io/Rdatasets/csv/Stat2Data/Election16.csv')
us2016_states =
  us2016_states[, result_state := fifelse(TrumpWin==1, 'Trump won', 'Clinton won')] %>%
  .[, .(state_abbr = Abr, result_state)]

## Merge and melt (reshape longer)
us2016 =
  us2016_counties[us2016_states, on='state_abbr'] %>%
  melt(id = c('fips', 'state_abbr'), variable.name = 'geo', value.name = 'result') %>%
  .[, geo := gsub('result_', '', geo)]


# Merge -------------------------------------------------------------------

us =
  nytc[us2016, on = 'fips', allow.cartesian = TRUE] %>%
  .[!is.na(date),
    lapply(.SD, sum, na.rm = TRUE), .SDcols = c('cases', 'deaths'),
    by = .(date, geo, result)]

## Get daily counts and percentages
setorder(us, geo, result, date)
us[ , ':=' (daily_cases = cases - shift(cases, 1, 'lag'),
            daily_deaths = deaths - shift(deaths, 1, 'lag')),
    by = .(geo, result)] %>%
  .[, ':=' (daily_cases_perc = daily_cases/sum(daily_cases),
            daily_deaths_perc = daily_deaths/sum(daily_deaths)),
    by = .(geo, date)]

## Some labeling sugar to match the WaPo graphic
us$geo = factor(us$geo, levels = c('state', 'county'), labels = c('States', 'Counties'))

# Plot(s) -----------------------------------------------------------------

theme_set(
  theme_ipsum(grid = 'Y') +
    theme(legend.title = element_blank(),
          legend.position = 'bottom',
          axis.title.x = element_blank(),
          axis.title.y = element_blank())
  )

## Cases
ggplot(us[date>'2020-03-01'], aes(date, daily_cases_perc, col = result)) +
  geom_line() +
  scale_color_brewer(palette = 'Set1', direction = -1) +
  scale_y_percent(limits = c(0,1)) +
  labs(title = 'Where new cases have been reported each day',
       caption = 'Data: NY Times\nBased on: https://wapo.st/2ZimmCA\nCode: https://github.com/grantmcdermott/covote') +
  facet_wrap(~ geo)

## Deaths
ggplot(us[date>'2020-03-01'], aes(date, daily_deaths_perc, col = result)) +
  geom_line() +
  scale_color_brewer(palette = 'Set1', direction = -1) +
  scale_y_percent(limits = c(0,1)) +
  labs(title = 'Where new deaths have been reported each day',
       caption = 'Data: NY Times\nBased on: https://wapo.st/2ZimmCA\nCode: https://github.com/grantmcdermott/covote') +
  facet_wrap(~ geo)
```

## Acknowledgements

- [Original WaPo article by Philip Bump](https://www.washingtonpost.com/politics/2020/06/17/coronavirus-has-come-trump-country/)
- [NY Times COVID-19 Data](https://github.com/nytimes/covid-19-data)
- [US 2016 election results by county (Tony McGovern)](https://github.com/tonmcg/US_County_Level_Election_Results_08-16)
- [US 2016 election results by state (Vincent Arel-Bundock)](https://vincentarelbundock.github.io/Rdatasets/)